### 2021-06-30 rad-icfp-2021/intro.tex
minors
writing appendix
simple writing phase
minor

\section{Introduction}

@@

The lack of purity of reverse-mode is known to make it significantly harder to optimise and find parrallelism. 
None of the current implementations of reverse-mode in a functional setting \cite{lantern_icfp,pearlmutter2008reverse,baydin2016diffsharp} are pure, and often complicated heuristics with no guarantees are used, e.g. \cite{xla}.
It means that for every important non trivial operation, a hand-crafted reverse-derivative must be given to be efficient.
In other words functional implementations seem to stuggle to abstract away from imperative code.

% For example, \MH{TODO: show how $\prod_i v_i$ for a vector $v$ is done in some functional AD systems}.
% For real world AD systems, modularity is key. They need to be expressive but have a lot of optimised derivatives for operators that are used a lot.
% This is easily seen on numpy \MH{TODO: explain a bit better why torch is good by using optimised grad for numpy.}
An efficient purely functional reverse-mode transformation would be more easily explainable, optimisable, and reusable.

@@

\subsection{Smmary of contributions}

@@

Our contribution tries to make these claims precise. Our detailed  contribution consist of
\begin{itemize}
    \item introducing an expressive source language with array operations, yet restrictive enough for efficient differentiation
    \item a method for extending our work and modular method for adding and prove correct new primitives as needed in practice 
    \item a new intermediate representation (I.R.) which we call unary form (UNF), essentially turning every operation to a unary one
    \item a simple reverse-mode transformation on the I.R.
    \item an expressive purely functional target language and efficient transformation from the I.R. to this language
    \item several useful generalisations in Section~\ref{sec:generalisation}. 
    \item correctness of the differentiation \ref{sec:correctness}, and complexity guarantees \ref{sec:complexity}
    \item showing how efficiency  is recovered via standard optimisation techniques
    \item a detailed comparison with other work and some limitations to our work in Section~\ref{sec:conclusion}
\end{itemize}

@@

 \begin{example}[Simple operations on arrays]
    On arrays, three simple operations of interest are the dot-product of two vectors, and the product or sum of the elements of a vector.
    In a functional setting, these can be defined as follows:
\begin{center}
    \begin{tabular}{{r c l}}
        !prod(A)! &:=& !reduce * 1 A! \\
        !sum(A)! &:=& !reduce + 0 A! \\
        !dot(A,B)! &:=& !reduce + 0 (map2 * A B)!     
    \end{tabular}
\end{center}
where !reduce! is a known fold left operator for which the function argument is associative and is faster than a fold left.

The gradient of each of these construct w.r.t. !A! will be:
\begin{center}
    \begin{tabular}{{r c l}}
        $\nabla_A$!prod(A)! &:=& !map2 * (scanr * 1 A) (shift1(scanl * 1 A))!\\
        $\nabla_A$!sum(A)! &:=& !map (x -> 1) A!\\
        $\nabla_A$!dot(A,B)! &:=& !map2 * B (map (x -> 1) A)! 
    \end{tabular}
\end{center}
where !shift1 [v1,...,vn] = [1,v1,...,v(n-1)]! is a simple operator yet perhaps strange at first naturally showing up in our constructions.
!scanl! and !scanr! are respectively scan left and scan right. These are like fold but store all intermediate results in an array.
We obtain these gradients, which are efficient and usually derived by hand, as special cases of our work.
They are obtained automatically in our case.
\end{example}

>>>>

\section{Introduction}
\label{sec:intro}

@@

The lack of purity of reverse-mode is known to make it significantly harder to optimise and find parallelism. 
None of the current implementations of reverse-mode in a functional setting \cite{lantern_icfp,pearlmutter2008reverse,baydin2016diffsharp} are pure, and often complicated heuristics with no guarantees are used, e.g. \cite{xla}.
It means that for every important non trivial operation, a hand-crafted reverse-derivative must be given to be efficient.
In other words functional implementations seem to struggle to abstract away from imperative code.

% For example, \MH{TODO: show how $\prod_i v_i$ for a vector $v$ is done in some functional AD systems}.
% For real world AD systems, modularity is key. They need to be expressive but have a lot of optimised derivatives for operators that are used a lot.
% This is easily seen on numpy \MH{TODO: explain a bit better why torch is good by using optimised grad for numpy.}
% An efficient purely functional reverse-mode transformation would be more easily explainable, optimisable, and potentially reusable.
 An efficient purely functional reverse-mode transformation would be more easily explainable, optimisable, and reusable.

@@

\subsection{Summary of contributions}

@@

Our contribution tries to make these claims precise. Our detailed  contribution consist of
\begin{itemize}
    \item introducing an expressive source language with array operations, yet restrictive enough for efficient differentiation
    \item a method for extending our work and modular method for adding and prove correct new primitives as needed in practice 
    \item a new intermediate representation (I.R.) which we call unary form(UNF), essentially turning every operation to a unary one
    \item a simple reverse-mode transformation on the I.R.
    \item an expressive purely functional target language and efficient transformation from the I.R. to this language
    \item several useful generalisations in Section~\ref{sec:generalisation}. 
    \item correctness of the differentiation \ref{sec:correctness}, and complexity guarantees \ref{sec:complexity}
    \item showing how efficiency  is recovered via standard optimisation techniques
    \item a detailed comparison with other work and some limitations to our work in Section~\ref{sec:conclusion}
\end{itemize}

@@

 \begin{example}[Simple operations on arrays]
    On arrays, three simple operations of interest are the dot-product of two vectors, and the product or sum of the elements of a vector.
    In a functional setting, these can be defined as follows:
\begin{center}
    \begin{tabular}{{r c l}}
        !prod(A)! &:=& !reduce * 1 A! \\
        !sum(A)! &:=& !reduce + 0 A! \\
        !dot(A,B)! &:=& !reduce + 0 (map2 * A B)!     
    \end{tabular}
\end{center}
where !reduce! is a known fold left operator for which the function argument is associative. 
It is notably faster to execute than a fold left as it parallel friendly.

The gradient of each of these construct w.r.t. !A! will be:
\begin{center}
    \begin{tabular}{{r c l}}
        $\nabla_A$!prod(A)! &:=& !map2 * (scanr * 1 A) (shift1(scanl * 1 A))!\\
        $\nabla_A$!sum(A)! &:=& !map (x -> 1) A!\\
        $\nabla_A$!dot(A,B)! &:=& !B! 
    \end{tabular}
\end{center}
where !shift1 [v1,...,vn] = [1,v1,...,v(n-1)]! is a simple operator yet perhaps strange at first naturally showing up in our constructions.
!scanl! and !scanr! are respectively scan left and scan right. These are like fold but store all intermediate results in an array.
We obtain these gradients, which are efficient and usually derived by hand, as special cases of our work.
They are obtained automatically in our case.
\end{example}
